### 2021-08-31 paper.Rmd
done with reviewer 2

\begin{figure*}
   \centering
         \includegraphics[width=0.9\textwidth]{figures/scheidigger_nn_tour.png}
\caption{A grand tour run on the output of a deep neural network model for classification. The axes correspond to the probabilities that the model of an image belonging to a certain class. Adapted from \cite{Li2020-kg}.}
\label{fig:tour-nn}
\end{figure*}

The grand tour along with direct interaction techniques has been proposed as a technique for understanding the training of deep neural network models [@Li2020-kg]. They propose a novel method for aligning the output of different layers by interpolating the output of different layers of a deep neural network via the the grand tour. They also apply the grand tour to explore the changes in single layers of the network for each training epoch. For example, by running the tour over the softmax layer outputs of a classification model, an analyst may understand where in the training confusion between classes occur and gain a view of the model performance (Figure \ref{fig:tour-nn}). Similarly, the difference between training and test sets can be assessed through side-by-side linked tours of layer outputs. @li_toward_2020 extends this idea by first reducing the dimensionality of output layers with UMAP [@mcinnes2020umap] and then using the grand tour with manual controls over 15 dimensional embeddings to understand layer topology.

@@

\begin{figure*}
  \centering
           \includegraphics[width=0.45\textwidth]{figures/allcenter-301.png}
         \includegraphics[width=0.45\textwidth]{figures/jetCluster-027.png}
\caption{Selected frames from using a grand tour of the physics experiments data. The three different types of measurements (shown in different colors) are aligned along different directions in parameter space (left). An outlying point is marked with an asterisk symbol (grey) in the ATLAS7new facet (right). Adapted from \cite{Cook2018-jm}, Figures 6 and 7.}
\label{fig:physics}
\end{figure*}

@@

In single cell RNA sequencing, scientists are interested in identifying novel cell types and understanding the relationships between cells or their developmental trajectory. To achieve this they perform cluster analysis on a counts matrix or principal components and embed the results via t-SNE and label points according the cluster label. One of the main advantages of t-SNE is the avoidance of over-plotting so clusters can be clearly identified on a scatter plot, however, this can come at the cost of interpretability as global distances are distorted.  In @Laa2020-uv, we used radial transformations of the tour projections as an alternative to t-SNE that better preserves global structure while still retaining cluster topology.

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.24\textwidth]{figures/torus-rings-4-with-torus-1.pdf}\includegraphics[width=0.24\textwidth]{figures/torus-rings-4-with-torus-2.pdf}
  \includegraphics[width=0.24\textwidth]{figures/torus-flat-6-1.pdf}\includegraphics[width=0.24\textwidth]{figures/torus-flat-6-2.pdf}
\caption{Several views of torii: (two at left) 4D ring torus, and (two at right) 6D flat torus. In each case, a subset has been highlighted which illustrate the low-dimensional donut shapes. Adapted from \cite{RJ-2016-044}, Figures 10 and 13.}
\label{fig:geozoo}
\end{figure*}

@@

This review highlights the history of tour methods, and modern extensions for the visualization of high dimensional numerical data. The rich ecosystem of tours is being actively developed and applied to a broad range of problems including machine learning. When tours are be combined with some interactivity it can assist with analyses like clustering (the spinifex and liminal packages).  Modified displays of projected data can be used to expose local structures (the sage algorithm, slicing and section pursuit). New algorithm data collection inside the tour code is being used to better diagnose optimization procedures for projection pursuit [@zhang2021visual].

There are many possible future directions for tour research.  One big challenge for software engineering is to seamlessly embed the tour with interactivity. Currently, there is no implementation that analysts can use to perform tasks that were easy in software available in the 80s and 90s. More contributions in sectioning as an alternative to projection, could be broadly useful. An example is conditional model visualization, as implemented in the condviz package [@JSSv081i05], which assists with visual exploration of multivariate fitted models. It could be interesting to define tour paths for different types of nonlinear subspaces, or non-Euclidean space.

>>>>

\begin{figure*}
   \centering
         \includegraphics[width=0.9\textwidth]{figures/scheidigger_nn_tour.png}
\caption{A grand tour run on the output of a deep neural network model for classification the benchmark clothing image data, fasion-MNIST. The axes correspond to the probabilities that the model of an image belonging to a certain class. Adapted from \cite{Li2020-kg}.}
\label{fig:tour-nn}
\end{figure*}

The grand tour along with direct interaction techniques has been proposed as a technique for understanding the training of deep neural network models [@Li2020-kg]. Using the benchmark data fashion-MNIST [@fashion-MNIST], they describe a novel method for aligning the output of different layers by interpolating the output of different layers of a deep neural network via the the grand tour. They also apply the grand tour to explore the changes in single layers of the network for each training epoch. For example, by running the tour over the softmax layer outputs of a classification model, an analyst may understand where in the training confusion between classes occur and gain a view of the model performance (Figure \ref{fig:tour-nn}). Similarly, the difference between training and test sets can be assessed through side-by-side linked tours of layer outputs. @li_toward_2020 extends this idea by first reducing the dimensionality of output layers with UMAP [@mcinnes2020umap] and then using the grand tour with manual controls over 15 dimensional embeddings to understand layer topology.

@@

\begin{figure*}
  \centering
           \includegraphics[width=0.45\textwidth]{figures/allcenter-301.png}
         \includegraphics[width=0.45\textwidth]{figures/jetCluster-027.png}
\caption{Selected frames from using a grand tour of the first six PCs of data on particle physics experiments. Each point is a physical observable and measures how the fit prediction changes in a 56D parameter space. The three different types of measurements (shown in different colors) are aligned along different directions in parameter space (left). An outlying point is marked with an asterisk symbol (grey) in the ATLAS7new facet (right). Adapted from \cite{Cook2018-jm}, Figures 6 and 7.}
\label{fig:physics}
\end{figure*}

@@

In single cell RNA sequencing, scientists are interested in identifying novel cell types and understanding the relationships between cells or their developmental trajectory. To achieve this, they perform cluster analysis on a counts matrix or principal components and embed the results via t-SNE and label points according the cluster label. One of the main advantages of t-SNE is the avoidance of over-plotting so clusters can be clearly identified on a scatter plot, however, this can come at the cost of interpretability as global distances are distorted.  @Laa2020-uv used radial transformations of the tour projections as an alternative to t-SNE that better preserves global structure while still retaining cluster topology (Figure \ref{fig:mouse-cluster}).

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.24\textwidth]{figures/torus-rings-4-with-torus-1.pdf}\includegraphics[width=0.24\textwidth]{figures/torus-rings-4-with-torus-2.pdf}
  \includegraphics[width=0.24\textwidth]{figures/torus-flat-6-1.pdf}\includegraphics[width=0.24\textwidth]{figures/torus-flat-6-2.pdf}
\caption{Several views of torii: (two at left) 4D ring torus, and (two at right) 6D flat torus. In each plot, the highlighted points illustrate the low-dimensional donut shapes that build the high-dimensional torus. Adapted from \cite{RJ-2016-044}, Figures 10 and 13.}
\label{fig:geozoo}
\end{figure*}

## Historical applications {#history}

Over the years there have been numerous applications of tours for problems solving. Here are a few examples. @Koschat-Swayne used the tour for telecommunication data to explore clustering of customers. @Symanzik2002NewAO used the image grand tour to detect locations of mines from satellite images, and traces of rock art beneath graffiti from photos. Census data is explored in @huh-kim. The multicollinearity between predictors in a large regression model on political preferences based on demographics is check with a tour by @forbes-cook-hyndman. Clustering is a common application for tours, and @newell-cook-hofmann-jannink discusses exploring clusters in population genetics of food crops. @macedo-cook-brown apply the grand tour to examine multiple climate variables in relation to their spatial coordinates.

@@

This review highlights the history of tour methods, and modern extensions for the visualization of high dimensional numerical data. The rich ecosystem of tours is being actively developed and applied to a broad range of problems including machine learning. When tours are combined with some interactivity supporting software can assist with analyses like clustering (the spinifex and liminal R packages).  Modified displays of projected data can be used to expose local structures (the sage algorithm, slicing, and section pursuit). New algorithm data collection inside the tour code is being used to better diagnose optimization procedures for projection pursuit [@zhang2021visual].

It needs to be stated that using tours can enhance our understanding of relationships between many variables, and typically leads to detecting previously unknown structure. Illustrations supporting this statement can be found in many of the applications cited in Section \@ref(history). A purpose of this review paper is to encourage broader use of tour methods.

There are many possible future directions for tour research.  One big challenge for software engineering is to seamlessly embed the tour with interactivity. Currently, there is no implementation that analysts can use to perform tasks that were easy to perform in software available in the 80s and 90s, particularly linking between dynamic tours and other displays (histograms, scatterplots or maps). More contributions in sectioning, as an alternative to projection, could be broadly useful. An example is conditional model visualization, as implemented in the condviz package [@JSSv081i05], which assists with visual exploration of multivariate fitted models. It could be interesting to define tour paths for different types of nonlinear subspaces, or non-Euclidean spaces.
